Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach

### 2022-03-10 analysis/paper/paper.Rmd
Updated text and overhauled certain figures

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].

@@

comb1 <-
  visnir.p1 + visnir.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
                              legend.background = element_rect(linetype = "solid", color = "black"),
                              legend.position = "bottom")

comb1 +
  plot_layout(nrow = 2, guides = "collect")
```

@@

```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra. The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands. As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA. PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the top left (marked with red circle), that may result from Al, Si and OH vibrations."}

@@

#Set up the score plot and 
raman.p2 <- 
  r.data2 %>%  
  ggplot(aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  labs(x = raman.pc1lab,
       y = raman.pc3lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5)) +
  #Filter out points with 700 - 800 wavenumber peaks and give them a red circle and slightly increase their size for better identification
  geom_point(data = r.data2 %>% 
               filter(sample_id == 6 | sample_id == 34 | sample_id == 42 | sample_id == 127 | sample_id == 445 | sample_id == 482 | sample_id == 491 | sample_id == 494 | sample_id == 495 | sample_id == 501 |sample_id == 502 | sample_id == 503 | sample_id == 504), 
             pch=21,
             size=4, 
             colour ="red") +
  theme(legend.position = "none")

comb <-
  raman.p1 + raman.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
                              legend.background = element_rect(linetype = "solid", color = "black"),
                              legend.position = "bottom")

comb +
  plot_layout(nrow = 2, guides = "collect")

@@

## XRF
Four samples had dimensions that were < 10 mm in size of their smallest dimension, and were therefore excluded from the analysis as there may be complications when comparing these to larger samples [@shackley11b pp.9-10]. The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal) that represents the remaining unquantifiable elements in a sample. The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF). Univariate analysis shows that SiO^2^ makes up the main bulk of the composition of the material, and two major groups can be distinguished based on this content (fig. \@ref(fig:xrf-violin-plot)). The majority of the readings resulted in a SiO^2^ content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO^2^ content closer to 90%. The “low” SiO^2^ group also feature a Bal between 40 – 70%.

The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 81.3%. The loadings can be be seen in figure \@ref(fig:xrf-loadings). In the score plot of PC 1 and PC 2 two large clusters can be seen (fig. \@ref(fig:xrf-score-plot)). Interestingly, the smaller cluster on the left side in the score plot includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content. It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that are the main contributors. The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom. This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711]. The brecciated quartz samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster. The white (N9) material is fairly evenly distributed within the right cluster.

The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river. The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %. This artefact was sampled twice on each side, resulting in similar results both times.

@@

#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
```

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of percentage element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 474 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
Points_long <-
  drop_na(Points.xrf %>%
    filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
    select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
    gather(key = "Element", value = "Value"))

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each variable (element)"}
#filter on sample size (only include samples with smallest dimension >= 10mm), 
#select elements to include in pca, as well as columns to group by
xrf <-
  Points.xrf %>%
  filter(max_length_mm >= 10 & max_width_mm >= 10) %>% 
  select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)

#log transform data
xrf.log <- log(xrf[,5:20])

@@

Although no distinct enough groupings were observed as to make classification possible, a number of structural and compositional characteristics were identified. Despite the Raman PCA lacking descriptive power, due to the few features observed in the spectra, the presence of the “graphite peaks” may present an important piece of information in the identification of the potential geological origin of the material. Interestingly, mentions of graphitic quartzite can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.

Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies are necessary in order to infer resource management strategies from this evidence. The characterization of quartz and quartzite is fundamental, however, if questions of resource procurement, trade and exchange and human mobility are to be developed. Being able to distinguish between different types of quartz and quartzite would enable various spatial analyses and modelling of distribution patterns, regardless of whether the geological source is known or not.  Moving forwards, the answer may lie in the field of multi-block analysis, where the potential for joint relations between separate blocks of data can be tested.

@@

The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification. Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials. The study represents a good example of a multivariate exploratory approach to the characterization of petrographic material where there also is an absence of reference material. There is a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources. Taking full advantage of the short integration time of these instrumentations it is possible to build large comprehensive datasets that then can be explored and mined using multivariate models.

>>>>

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].   
 \linebreak

@@

#Set the legend for the combined figure of both score plots
comb1 <-
  visnir.p1 + visnir.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
                              legend.background = element_rect(linetype = "solid", color = "black"),
                              legend.position = "bottom")

#Call on the finished figure and layout in 2 rows
comb1 +
  plot_layout(nrow = 2, guides = "collect")
```

@@

```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra. The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands. As these are the main features in the data, the other components mainly describe the remaining noise. PC 3 does, however, separate out the few samples that feature peaks at 700-800 cm^-1^ in the top left (marked with red circle), that may result from Al, Si and OH vibrations."}

@@

#Set up the score plot
raman.p2 <- 
  r.data2 %>%  
  ggplot(aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  labs(x = raman.pc1lab,
       y = raman.pc3lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5)) +
  #Filter out points with 700 - 800 wavenumber peaks and give them a red circle for easier identification
  geom_point(data = r.data2 %>% 
               filter(sample_id == 6 | sample_id == 34 | sample_id == 42 | sample_id == 127 | sample_id == 445 | sample_id == 482 | sample_id == 491 | sample_id == 494 | sample_id == 495 | sample_id == 501 |sample_id == 502 | sample_id == 503 | sample_id == 504), 
             pch=21,
             size=7, 
             colour ="red",
             fill = NA,
             stroke = 2) +
  theme(legend.position = "none")

#Set the legend for the combined figure of both score plots
comb <-
  raman.p1 + raman.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
                              legend.background = element_rect(linetype = "solid", color = "black"),
                              legend.position = "bottom")

#Call on the finished figure and layout in 2 rows
comb +
  plot_layout(nrow = 2, guides = "collect")
```

@@

@@

#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"

#Filter out Si values above 50%
high_si <- 
  Points.xrf %>% 
    filter(`14Si` > 50)

#Prepare a calculation of percentage measurements above 50% to call on in text
high_si <- 
  round(length(high_si$`14Si`) / length(Points.xrf$`14Si`) * 100)

#Prepare XRF data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
Points_long <-
  drop_na(Points.xrf %>%
    filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
    select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
    gather(key = "Element", value = "Value"))

#filter XRF data on sample size (only include samples with smallest dimension >= 10mm), 
#select elements to include in PCA, as well as columns to group by
xrf <-
  Points.xrf %>%
  filter(max_length_mm >= 10 & max_width_mm >= 10) %>% 
  select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)

#log transform XRF data
xrf.log <- log(xrf[,5:20])

#Prepare cumulative explained variance of two first XRF PCs
xrf_variance <- 
  round(summary(xrf.pca)$importance[2,1]*100, digits=1) + round(summary(xrf.pca)$importance[2,2]*100, digits=1)

## XRF
Four samples had dimensions that were < 10 mm in size of their smallest dimension, and were therefore excluded from the analysis as there may be complications when comparing these to larger samples [@shackley11b pp.9-10]. The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal) that represents the remaining unquantifiable elements in a sample. The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF). Univariate analysis shows that ^14^Si makes up the main bulk of the composition of the material, with a bimodal distribution (fig. \@ref(fig:xrf-violin-plot)). The majority of the readings resulted in a ^14^Si content of less than 50%, and about ¼ of the readings (~ `r high_si`%) featured a ^14^Si content closer to 90%. The “low” ^14^Si group also feature a Bal between 40 – 70%.

The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ `r high_si`%. The loadings can be be seen in figure \@ref(fig:xrf-loadings). In the score plot of PC 1 and PC 2 two large clusters can be seen (fig. \@ref(fig:xrf-score-plot)). The smaller cluster on the left side in the score plot includes the samples with a ^14^Si content higher than 50%, whilst the larger cluster includes those with lower ^14^Si content. It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that are the main contributors. The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom. This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711]. The brecciated quartz samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster. The white (N9) material is fairly evenly distributed within the right cluster.

The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river. The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25%. This artefact was sampled twice on each side, resulting in similar results both times.

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of percentage element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 474 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each variable (element)"}

@@

Although no distinct enough groupings were observed as to make classification possible, a number of structural and chemical characteristics were identified. Despite the Raman PCA lacking descriptive power, due to the few features observed in the spectra, the presence of the “graphite peaks” may present an important piece of information in the identification of the potential geological origin of the material. Interestingly, mentions of graphitic quartzite can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day. Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies would be necessary in order to infer resource management strategies from this evidence.

The exploratory approach presented in this paper holds potential for enabling provenance studies within geologically complex regions. Screening a large enough dataset makes it possible to look at in-group variations and isolate spectroscopic distinct fingerprints [e.g. @sciuto_etal19]. These can then be used as the basis for later classification of the material into structurally different groups. The characterization of quartz and quartzite is fundamental if questions of resource procurement, trade and exchange and human mobility are to be developed within Northern Sweden. Being able to distinguish between different types of quartz and quartzite would enable the spatial analysis and modelling of distribution patterns, regardless of whether the geological source is known or not. In the current case study modelling the results of single instrumentation may have been limited for the purpose of classification, but modelling the joint relations between the separate data blocks may provide further aid in isolating distinct types of quartz and quartzite.

@@

The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification. Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials. The study represents a good example of a multivariate exploratory approach to the characterization of petrographic material where there also is an absence of reference material. There is a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources. Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined using multivariate models.
